Stage-Splitting a Modal Language

### 2014-09-05 paper/semantics.tex
Minor grammatical improvements for clarity.

When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  We call this search process
\emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes
a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
speculated into a residual ($q$), we then return a fresh variable wrapped in a
$\next$ block ($\next~\hat y$), along with a future environment which maps that
variable to our residual ($\hat y \mapsto q$).

At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
nothing but recursively speculate into every subexpression.  Once speculation
finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which
produces a future context and (by canonical forms) a $\next$-wrapped variable
($\next\{\hat y\}$)\footnote{
Importantly, speculation will search into both branches of a stage-\bbtwo\ {\tt if}
statement to evaluate the contained stage-\bbone\ code.  Thus the name.
}.  The context is then reified (using the $\reifysym$
judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of
its $\next$---at the bottom.

@@

If we change the $\next$ and $\prev$ rules to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value, where $q$ is a residual,
then we get precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
our semantics terminates strictly less often than that of \cite{davies96}.

@@

The dynamics described in the previous section provide a formal description of partial evaluation.
Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is residual, also called ``f specialized to x."

>>>>

When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  We call this search process
\emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes
a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
speculated into a residual ($q$), we then return a fresh variable wrapped in a
$\next$ block ($\next~\hat y$), along with a future environment which maps that
variable to our residual ($\hat y \mapsto q$).

At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
nothing but recursively speculate into every subexpression.  Once speculation
finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which
produces a future context and (by canonical forms) a $\next$-wrapped variable
($\next\{\hat y\}$)\footnote{
Importantly, speculation will search into both branches of a stage-\bbtwo\ {\tt if}
statement to evaluate all contained stage-\bbone\ code.  Thus the name.
}.  The context is then reified (using the $\reifysym$
judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of
its $\next$---at the bottom.

@@

If we change the $\next$ and $\prev$ rules to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value, where $q$ is a residual,
then we get precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

@@

The dynamics described in the previous section provide a formal description of partial evaluation.
Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."

### 2014-09-05 paper/splitting.tex
Minor grammatical improvements for clarity.

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before any arguments are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

@@

Recall that, along with a future environment, partial values are the output of the partial evaluation of stage \bbone\ terms.
In addition to base types, tuples, and injections, partial values may also contain references into the future environment.
In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
The goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 
and $v_2$ contains only the information from $v$ that is likewise relevant to second stage code.
We call this action {\em masking}.

>>>>

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before {\em any arguments} are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

@@

Recall that, along with a future environment, partial values are the output of the partial evaluation of stage \bbone\ terms.
In addition to base types, tuples, and injections, partial values may also contain references into the future environment.
In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
The goal of partial value splitting is to take a partial value $v$ and from it create two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 
and $v_2$ contains only the information from $v$ that is likewise relevant to second stage code.
We call this action {\em masking}.
